A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-02 02-ceu.tex
Revisão 2.1.1: - adicionado event iterator "every", vai ser importante para forçar terminação

|\C{1.}{Control flow}|
$\T<stmt>$ ; $\T<stmt>$                       | \C{.45}{sequence}     |
if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end | \C{.45}{conditional}  |
loop do $\T<stmts>$ end                   | \C{.45}{repetition}   |
finalize [$\T<stmts>$] with $\T<stmts>$ end    | \C{.45}{finalization} |

@@

In \CEU, the queue-based processing of events described previously applies
only to external input events, i.e., those events submitted to the program
by the environment.
%
Internal events, which are those events generated internally by the program
via \code{emit} statements, are processed in a stack-based manner.
%
These internal events provide a fine-grained execution control and, because
of their stack-based processing, can be used to implement a limited form of
subroutine, as illustrated in Listing~\ref{lst.sub} below.

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A \CEU program with a ``subroutine''.},
  label={lst.sub},
]
event int* inc;     // subroutine |`|inc|'|
par/or do
    loop do         // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    $\cdots$
    emit inc(&v);   // call |`|inc|'|
    _assert(v==2);  // after return
end
\end{lstlisting}
\vskip-\baselineskip

In Listing~\ref{lst.sub}, the ``subroutine'' \code{inc} is defined as a loop
(ln.~3--6) that continuously awaits its identifying event (ln.~4), and
increments the value passed to it by reference (ln.~5).
%
A trail in parallel (ln.~8--11) invokes the subroutine through an \code{emit}
statement (ln.~10).
%
Given the stack-based execution mode of internal events, as the emit executes,
the calling trail pauses, the subroutine awakes (ln.~4), runs its body
(yielding \code{v=2}), loops, and awaits the next ``call'' (ln.~4, again).
%
Only after this sequence does the calling trail resumes and moves on to
execute the assertion (ln.~11).

\CEU also supports nested \code{emit} invocations for internal events.
%
For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
could \code{emit} another internal event after awaking (ln.~4), creating a
new level in the stack.
%
We can think of the stack as a record of the nested, fine-grained internal
reactions that happen inside the same bigger reaction to some external
event.

This form of subroutine has a significant limitation though: it cannot
express recursion (an \code{emit} to itself is always ignored as a running
trail cannot be waiting on itself).
%
That said, it is this very limitation that brings important safety
properties to subroutines.
%
First, such subroutines are guaranteed to react in bounded time.
%
Second, memory for locals is also bounded, not requiring data stacks.
%
Third, \CEU subroutines can be safely used by the other primitives of the
language, such as parallel compositions and the \code{await} statement,
without breaking the programming model.
%
In particular, after calling a subroutine these primitives wait while
keeping context information, such as locals and the program counter, which
makes the calls behave similarly to those of
coroutines~\cite{lua.coroutines}.
%
Finally, in previous work, we built other advanced control mechanisms on top
of internal events, such as resumable exceptions and reactive
variables~\cite{ceu.rem13}.
%
%In Section~\ref{sec.adv.excpt} we show how to use them to implement
%exceptions.
%
%8<- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\gl{Revisei do início da Seção~2 até aqui.}
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - ->8

>>>>

|\C{1.}{Control flow}|
$\T<stmt>$ ; $\T<stmt>$                       | \C{.45}{sequence}     |
if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end | \C{.45}{conditional}  |
loop do $\T<stmts>$ end                   | \C{.45}{repetition}   |
every $\T<id>$ in $\T<id>$ do $\T<stmts>$ end      | \C{.45}{event iteration}   |
finalize [$\T<stmts>$] with $\T<stmts>$ end    | \C{.45}{finalization} |

@@

In \CEU, queue-based processing of events applies only to external input
events, i.e., events submitted to the program by the environment.
%
Internal events, which are events generated internally by the program via
\code{emit} statements, are processed in a stack-based manner.
%
Internal events provide a fine-grained execution control, and, because of their
stack-based processing, can be used to implement a limited form of subroutines,
as illustrated in Listing~\ref{lst.sub}:

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A \CEU program with a ``subroutine''.},
  label={lst.sub},
]
event int* inc;         // declares subroutine "inc"
par/or do
    var int* p;
    every p in inc do   // implements subroutine with an event iterator
        *p = *p + 1;
    end
with
    var int v = 1;
    emit inc(&v);       // calls "inc"
    emit inc(&v);       // calls "inc"
    _assert(v==3);      // asserts result after returns
end
\end{lstlisting}
\vskip-\baselineskip

In the example, the ``subroutine'' \code{inc} is defined as an event iterator
(ln.~4--6) that continuously awaits its identifying event (ln.~4), and
increments the value passed by reference (ln.~5).
%
A trail in parallel (ln.~8--11) invokes the subroutine through two consecutive
\code{emit} statements (ln.~9--10).
%
Given the stack-based execution for internal events, as the first emit
executes, the calling trail pauses (ln.~9), the subroutine awakes (ln.~4), runs
its body (yielding \code{v=2}), iterates, and awaits the next ``call'' (ln.~4,
again).
%
Only after this sequence does the calling trail resumes (ln.~9), makes a new
invocation (ln.~10), and passes the assertion test (ln.~11).

%- multiple executions
%- footnote to distinguish from TECS

\CEU also supports nested \code{emit} invocations for internal events.
%
For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
could \code{emit} another internal event after awaking (ln.~4), creating a
new level in the stack.
%
We can think of the stack as a record of the nested, fine-grained internal
reactions that happen inside the same outer reaction to a single external
event.

This form of subroutines has a significant limitation though: it cannot
express recursion, since an \code{emit} to itself is always ignored as a
running trail cannot be waiting on itself.
%
That being said, it is this very limitation that brings important safety
properties to subroutines.
%
First, they guaranteed to react in bounded time.
%
Second, memory for locals is also bounded, not requiring data stacks.

At first sight, an event iterator for event \code{A} seems like a syntactic
sugar for a \code{loop} followed by and \code{await A}.
However, event iterators have an important restriction to enforce reaction
termination, as we discuss in Section~\ref{sec.sem}:
    they cannot contain \code{break} statements.

\begin{comment}
%
Third, \CEU subroutines can be safely used by the other primitives of the
language, such as parallel compositions and the \code{await} statement,
without breaking the programming model.
%
In particular, after calling a subroutine these primitives wait while
keeping context information, such as locals and the program counter, which
makes the calls behave similarly to those of
coroutines~\cite{lua.coroutines}.
%
Finally, in previous work, we built other advanced control mechanisms on top
of internal events, such as resumable exceptions and reactive
variables~\cite{ceu.rem13}.
%
\end{comment}
